---
output: cynkratemplate::readme_document
---

```{r setup, include = FALSE}
# README.md and .github/README.md are both written from this file, and differ
# by the banner alone. GitHub reads .github/README.md in preference to the
# root one, so that is where the wordmark goes; pkgdown reads the root one and
# already puts the package logo in its own header, where a second logo is one
# too many.
#
# The banner sits in .github/, one directory down, so its paths climb out.
knitr::knit_hooks$set(document = function(x, options) {
  doc <- paste(x, collapse = "\n")

  front <- strsplit(doc, "<!-- banner:start -->\n", fixed = TRUE)[[1]]
  rest <- strsplit(front[[2]], "<!-- banner:end -->\n\n", fixed = TRUE)[[1]]
  if (length(front) != 2L || length(rest) != 2L) {
    stop("README.Rmd: expected one banner:start and one banner:end marker")
  }

  # front[[1]] is the YAML, rest[[1]] the banner, rest[[2]] the shared body.
  # The blank line between them is load-bearing: without it the heading below
  # is swallowed into the banner's HTML block instead of being a heading.
  writeLines(paste0(rest[[1]], "\n", rest[[2]]), ".github/README.md")
  paste0(sub("\n+$", "\n", front[[1]]), rest[[2]])
})
```

<!-- banner:start -->
<div align="center">
  <a href="https://r.duckdb.org/"><picture>
    <source media="(prefers-color-scheme: light)" srcset="../man/figures/DuckDB_Logo-horizontal.svg">
    <source media="(prefers-color-scheme: dark)" srcset="../man/figures/DuckDB_Logo-horizontal-dark-mode.svg">
    <img alt="DuckDB logo" src="../man/figures/DuckDB_Logo-horizontal.svg" height="100">
  </picture></a>
</div>
<!-- banner:end -->

# duckdb

[DuckDB](https://duckdb.org/) is an in-process SQL OLAP database management system.
It is designed to support analytical query workloads and is optimized for fast query execution.
This repository contains the R bindings for DuckDB.

## Installation from CRAN
This is the recommended method for recent R versions on Windows or macOS which have binaries available on CRAN.

``` r
install.packages("duckdb")
```

For Linux or older R versions, installing the package from source may take up to an hour.
Consider the [Posit Public Package Manager](https://p3m.dev/) or [r-universe](https://duckdb.r-universe.dev) for binary installs (see the next sections).

## Installation from the Posit Public Package Manager

This repository serves binary builds of CRAN packages for a wide variety of platforms.
Follow setup instructions from <https://p3m.dev/client/#/repos/cran/setup>.
For example, to install the ManyLinux version of duckdb that is likely to work on your Linux, use:

``` r
install.packages("duckdb", repos = sprintf(
  "https://p3m.dev/cran/latest/bin/linux/manylinux_2_28-%s/%s", R.version["arch"], substr(getRversion(), 1, 3)
))
```

To check the availability of binaries for your platform, navigate to the [duckdb search page](https://p3m.dev/client/#/repos/cran/packages/overview?search=duckdb).

## Installation from r-universe

This repository serves development versions.
Binaries are available for recent versions of R and for some platforms.
Review <https://docs.r-universe.dev/install/binaries.html> for configuring installation of binary packages on Linux.

``` r
install.packages("duckdb", repos = c("https://duckdb.r-universe.dev", "https://cloud.r-project.org"))
```

Installing the package from source may take up to an hour.

## Installation from GitHub

``` r
# install.packages("pak", repos = sprintf("https://r-lib.github.io/p/pak/stable/%s/%s/%s", .Platform$pkgType, R.Version()$os, R.Version()$arch))
pak::pak("duckdb/duckdb-r")
```

Installing the package from GitHub may take up to an hour.

## Flavors

The sources in this repository are published under several names —
`duckdb` on CRAN,
and every flavor on [r-universe](https://duckdb.r-universe.dev/builds).
The `.dev` flavors are created by an automated vendoring process;
the CRAN and LTS flavors always point at a stable upstream release.

```{r flavors, echo = FALSE, results = "asis"}
# The table and the archive list below it are written from scripts/series.yaml;
# edit the declaration, not this file. scripts/series-table.R is the generator,
# and the handbook's copy of the same table is spliced by running it.
source("scripts/series-table.R")
decl <- series_decl(".")
cat(flavor_table(decl), "\n", sep = "")
```

Every row opens with the versions published, one badge per place it is
published from, and the flavor's name is a link on those rather than on the name
itself. Only `duckdb` is in two: CRAN releases it, and r-universe rebuilds the
same source between releases.
The three after it track a `.dev` series:
*ahead* counts commits ahead of the branch the series releases from,
*in flight* counts commits in CI but not yet trusted,
*buffered* counts commits vendored but not yet verified.
*ahead* is counted here, where the branch a series releases from lives and
`<S>-green` is mirrored to; the other two are counted in the fork, which is the
only repository carrying `<S>-dev` and `<S>-build`.
So the two greens either side of *in flight* are the same commit because one
push writes both and the loop stops if the second cannot fast-forward.

```{r archived, echo = FALSE, results = "asis"}
cat(archived_list(decl), "\n", sep = "")
```

See the handbook's
[`branches/`](https://github.com/duckdb/duckdb-r/blob/main/handbook/branches/README.md)
for the full model.

## User Guide

See the [R API in the DuckDB documentation](https://duckdb.org/docs/api/r).

## Documentation

Everything this repository documents is reachable from the
[handbook](https://github.com/duckdb/duckdb-r/blob/main/handbook/README.md),
a strict topic hierarchy;
[`AGENTS.md`](https://github.com/duckdb/duckdb-r/blob/main/AGENTS.md)
is the door for maintainers and coding agents, and
[`plan/README.md`](https://github.com/duckdb/duckdb-r/blob/main/plan/README.md)
names the designs, plans, and historical documents.
[`BRANCHES.md`](https://github.com/duckdb/duckdb-r/blob/main/BRANCHES.md) and
[`scripts/VENDORING.md`](https://github.com/duckdb/duckdb-r/blob/main/scripts/VENDORING.md)
hold detail the handbook is still absorbing.

## Building

To build the R package, you first need to clone this repository and install the dependencies:

``` r
# install.packages("pak", repos = sprintf("https://r-lib.github.io/p/pak/stable/%s/%s/%s", .Platform$pkgType, R.Version()$os, R.Version()$arch))
pak::pak()
```

Then, install:

``` sh
~duckdb-r: R CMD INSTALL .
```

Set the `MAKEFLAGS` environment variable to `-j8` or similar for parallel builds.
Configure `ccache` for faster repeated builds.
A build that links a prebuilt engine and finishes in seconds,
and other details, are described in the handbook under
[`build/`](https://github.com/duckdb/duckdb-r/blob/main/handbook/build/README.md).

## Vendoring

This package includes a vendored copy of the DuckDB C++ library. The vendoring process is automated — a scheduled routine synchronizes each release series with the upstream DuckDB repository, commit by commit. How it works, and how to vendor by hand, is in the handbook under [`operations/vendoring/`](https://github.com/duckdb/duckdb-r/blob/main/handbook/operations/vendoring/README.md).

## Contributors

Thanks to all [contributors](https://github.com/duckdb/duckdb-r/graphs/contributors) to this repository, and to those who contributed when the code was still hosted in the main [duckdb/duckdb](https://github.com/duckdb/duckdb) repository:

Mark Raasveldt, Pedro Holanda, Tom Ebergen, Reijo Sund, Nicolas Bennett, Patrik Schratz, Tishj, Laurens Kuiper, Sam Ansmink, Andy Teucher, Hadley Wickham, Jonathan Keane, Lindsay Wray, Richard Wesley, Elliana May, Edwin de Jonge, Dewey Dunnington, Carlo Piovesan, Andre Beckedorf, Tania Bogatsch, Pedro Ferreira, Maximilian Girlich, James Lamb, James Atkins, usurai, Ubuntu, Noam Ross, Michael Antonov, Jeroen Ooms, Jamie Lentin, Jacob, and Chilarai.
