---
title: "Memory-efficient reading and writing"
---

<!-- Handbook: handbook/usage/memory/README.md.
     This article is the user-facing digest of the memory leaves: the leaves own the facts, this page shows one way. -->

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

A result that fits comfortably in the database can still be too large for R,
and a data frame that fits in R can still be copied on its way into the database.
This article shows memory-efficient ways for reading and writing:

- Memory limits
- Streamed reading via Arrow
- Registering a data frame for writing

## Set the limit where it takes effect

DuckDB and R allocate from two different budgets.
The engine's memory is bounded by `memory_limit` and can spill to disk; R's vectors are unbounded and only spilled to the system swap space, if at all.
The routes below keep R's share to one batch, or one data frame, at a time, and leave the rest to the engine's budget.
The overheads quoted are measurements, described at the end of the page.

The engine's limit is set on the driver or on a live connection.

```{r}
library(DBI)
con <- dbConnect(duckdb::duckdb(config = list(memory_limit = "2GB")))
# or, on a live connection:
dbExecute(con, "SET memory_limit = '2GB'")
```

## Reading: stream, and release each batch

Reading via Arrow stream never holds the result whole.
Execution starts at `dbSendQueryArrow()` and pauses at a bounded buffer.
Each `dbFetchArrowChunk()` hands over one batch that can be converted using `as.data.frame()`.
A batch is released immediately via `nanoarrow_pointer_release()` or indirectly via `gc()` (assuming nothing else refers to it).

```{r}
rs <- dbSendQueryArrow(con, "SELECT * FROM huge")
repeat {
  batch <- dbFetchArrowChunk(rs, chunk_size = 1e6)
  if (batch$length == 0) break
  df <- as.data.frame(batch)
  # ... consume df ...
  nanoarrow::nanoarrow_pointer_release(batch)
}
dbClearResult(rs)
```

Overhead, measured: the loop above peaks between 160 and 230 MB over the process baseline for both an 800 MB and a 12 GB result on a 16 GB machine.

## Writing: hand the data frame over, and let the engine scan it

`dbWriteTable()` copies nothing on the R side:
it registers the data frame as a view that the engine scans in place, creates the table from that scan, and unregisters.
`dbAppendTable()` does the same into an existing table, so data that arrives in pieces is appended piece by piece with the same cost.

```{r}
dbWriteTable(con, "t", df)        # one frame
dbAppendTable(con, "t", next_df)  # or piece by piece
```

Overhead, measured: writing an 800 MB frame consumes between 280 and 320 MB in addition to the frame's own size,
writing a 12 GB frame on a 16 GB machine consumes 373 MB.
Where no table is needed at all, `duckdb::duckdb_register()` alone lets queries scan the frame for a measured 1 MB of overhead.

```{r}
duckdb::duckdb_register(con, "df", df)
dbGetQuery(con, "SELECT count(*) FROM df WHERE a > 0.5")
```

### Files: let the engine read them

A file the engine can read is cheapest read by the engine.
Calling `read_parquet()` function in the engine consumes 308 MB.

```{r}
dbExecute(con, "CREATE TABLE t AS SELECT * FROM read_parquet('data.parquet')")
```

### Arrow streams: append a batch at a time

Data that arrives as an Arrow stream, from a pipe, a socket or another library's reader,
goes in through `dbWriteTableArrow()`, or `dbAppendTableArrow()` into an existing table.
Both pull one batch on R's thread and append it as a data frame.
Measured: 505 MB over an empty process for the same 800 MB.

```{r}
stream <- nanoarrow::read_nanoarrow(file("data.arrows", "rb"))
dbWriteTableArrow(con, "t", stream)
```

## How the numbers were measured

Every number above is the peak resident memory of the whole R process over its baseline,
taken on the development build in September 2026, for the same 800 MB result or frame: fifty million rows of two `DOUBLE` columns.
The 12 GB figures are the same measurements at 750 million rows on a machine with 16 GB, one run at a time.
The method, every other route, and the same measurements through the Python, Node, Go and Rust clients are recorded in the repository:
[`experiments/2026-09-14-memory-clients/`](https://github.com/duckdb/duckdb-r/tree/main/experiments/2026-09-14-memory-clients) for reading,
[`experiments/2026-09-19-memory-ingest/`](https://github.com/duckdb/duckdb-r/tree/main/experiments/2026-09-19-memory-ingest) for writing.
