---
title: "Memory-efficient reading and writing"
---

<!-- Handbook: handbook/usage/memory/README.md — this article is the user-facing
     digest of the memory leaves; the leaves own the facts, this page shows one way. -->

```{r, include = FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

DuckDB and R allocate from two different budgets.
The engine's memory is bounded by `memory_limit` and can spill to disk;
R's vectors are bounded by nothing and never spill.
Every number below is a measurement of the same 800 MB result —
fifty million rows of two `DOUBLE` columns —
taken on the development build in September 2026,
as the peak resident memory of the whole R process over its baseline;
the method and every other route are recorded in the repository's
[`experiments/2026-09-14-memory-clients/`](https://github.com/duckdb/duckdb-r/tree/main/experiments/2026-09-14-memory-clients)
and
[`experiments/2026-09-19-memory-ingest/`](https://github.com/duckdb/duckdb-r/tree/main/experiments/2026-09-19-memory-ingest).

## Set the limit where it takes effect

The engine's limit is set on the driver, or on a live connection.
Passing it to `dbConnect(config = )` does not reach an instance that
already exists, which the default in-memory database always does.

```{r}
library(DBI)
con <- dbConnect(duckdb::duckdb(config = list(memory_limit = "2GB")))
# or, on a live connection:
dbExecute(con, "SET memory_limit = '2GB'")
```

## Reading: stream, and release each batch

The one route that never holds the result whole is the Arrow stream.
Execution starts at `dbSendQueryArrow()` and pauses at a bounded
buffer; each `dbFetchArrowChunk()` hands over one batch,
which `as.data.frame()` converts and `nanoarrow_pointer_release()`
frees at once.
The release is what gives you control: a batch's memory is Arrow's,
allocated by the engine outside R's heap, and R's garbage collector
never sees it, so a batch that is merely dropped waits for a
collection that nothing prompts.
`gc()` is the fallback, and it frees a batch only if nothing still
refers to it.

```{r}
rs <- dbSendQueryArrow(con, "SELECT * FROM huge")
repeat {
  batch <- dbFetchArrowChunk(rs, chunk_size = 1e6)
  if (batch$length == 0) break
  df <- as.data.frame(batch)
  # ... consume df ...
  nanoarrow::nanoarrow_pointer_release(batch)
}
dbClearResult(rs)
```

Overhead, measured: the loop above peaks between 160 and 230 MB over
the process baseline for the 800 MB result, and stays there even when
the query sorts more than `memory_limit` allows, because the sort spills
and the stream carries the result batch by batch.
It peaks at the same 166 to 229 MB for a 12 GB result on a 16 GB
machine: the peak belongs to the batch, not to the result.
Fetching the same result with `dbGetQuery()` peaks at about 1,650 MB,
twice the data: the engine's copy and R's coexist while the result
converts — and `dbSendQuery()` with a `dbFetch(n = )` loop peaks the
same, because the whole result is materialized before the first `n`
rows return.
At 12 GB, both are killed.
Dropping batches without releasing them costs about 900 MB,
the whole result waiting for a collector that runs on R-heap pressure it
never sees — and 8 GB of the 12 GB result before that collector
happens to run.
One thing the release does not do: a character column converted
from a batch keeps its part of the batch alive until the strings are
materialized or the column is dropped, because the conversion is lazy.
Numeric columns are copied, and the batch can go as soon as they are.

## Writing: hand the data frame over, and let the engine scan it

`dbWriteTable()` copies nothing on the R side: it registers the data
frame as a view the engine scans in place, creates the table from that
scan, and unregisters.
`dbAppendTable()` does the same into an existing table, so data that
arrives in pieces is appended piece by piece with the same cost.

```{r}
dbWriteTable(con, "t", df)        # one frame
dbAppendTable(con, "t", next_df)  # or piece by piece
```

Overhead, measured: writing an 800 MB frame peaks between 280 and
320 MB over the frame's own size, in a file database and in an
in-memory one alike under a 300 MB `memory_limit` — the engine's write
working set, bounded by the limit, and nothing else;
a 12 GB frame on a 16 GB machine peaks 373 MB over its own size.
Fifty appends of a million rows each, from a producer that never holds
the data whole, peak at about 430 MB over an empty process,
and 460 MB when the pieces add up to 12 GB.
Where no table is needed at all, `duckdb::duckdb_register()` alone
lets queries scan the frame for a measured 1 MB of overhead.

```{r}
duckdb::duckdb_register(con, "df", df)
dbGetQuery(con, "SELECT count(*) FROM df WHERE a > 0.5")
```

### Files: let the engine read them

A file the engine can read is cheapest read by the engine.
Routing a Parquet file through the arrow package's dataset scanner —
`arrow::open_dataset()` handed to `duckdb_register_arrow()`, or
`arrow::to_duckdb()` — buffers the file ahead of the table write and
peaks above the data's own size (measured: 1,036 MB for an 800 MB file),
where `read_parquet()` in the engine peaks at 308 MB, the pool at its
limit.

```{r}
dbExecute(con, "CREATE TABLE t AS SELECT * FROM read_parquet('data.parquet')")
```

### Arrow streams: append a batch at a time

Data that arrives as an Arrow stream — from a pipe, a socket, another
library's reader — goes in through `dbWriteTableArrow()`, or
`dbAppendTableArrow()` into an existing table: both pull one batch on
R's thread and append it as a data frame.
Measured: 505 MB over an empty process for the same 800 MB.
Registering the stream as a table instead costs the whole dataset for
the same reason the file does — about 2 GB once the data is larger than
that — and a stream over an R connection cannot be registered at all:
the engine pulls it from a thread that is not R's.

```{r}
stream <- nanoarrow::read_nanoarrow(file("data.arrows", "rb"))
dbWriteTableArrow(con, "t", stream)
```
